The ChEMBL database as Linked Open Data

### 2012-04-16 article.tex
more edits

Although many parts of the CHEMBL-RDF 13 dataset now use community based ontologies, some 
terms from the previous ad-hoc ontology are still used. These terms were created under the 
URI namespace \textit{http://rdf.farmbio.uu.se/chembl/onto/\#}, and are referenced here using the
prefix: \textit{chembl}.

@@

The core concept in the ChEMBL database is that of the biological activity. This
is the type of information that the ChEMBL database extracts from literature.
In CHEMBL-RDF activities are mostly represented using the custom CHEMBL-RDF vocabulary, where the
type, fields and links to other resources are using the \textit{chembl}
namespace. While ChEMBL provides both the original activities as found in literature
and standardized values allowing comparison between studies, the triples only
make the latter available. The CiTO ontology is used to point to the paper from
which the data was extracted.

@@

\begin{footnotesize}
\begin{verbatim}
mol:m41
 rdfs:label "CHEMBL406142" , 
   "Bis(3-[1 .... yl]-propionamide)" ;
 rdfs:subClassOf cheminf:CHEMINF_000000 ;
 cheminf:CHEMINF_000200
  m41:inchikey , m41:smiles , m41:inchi ;

@@

chemblid:CHEMBL406142
 owl:equivalentClass mol:m41 .
\end{verbatim}
\end{footnotesize}

@@

To make the data a proper citizen of the semantic web, we link out to various resources,
which is visualized in Figure~\ref{2}. Triples for compounds link out to ChemSpider 
using the new complementary index and OpenMolecules RDF using InChI values, to Bio2RDF
for protein targets using the uniprot identifier, and CrossRef and Bio2RDF for literature
references using the DOIs and PubMed identifiers, respectively.

@@

For protein links out are based on the FOOBAR for the species, the EC code for proteins,
as well as UniProt identifiers.

FIXME: should we link out to the new UniProt RDF?

@@

\begin{tiny}
\begin{verbatim}
<http://data.kasabi.com/dataset/chembl-rdf/chemblid/CHEMBL324846>
 skos:exactMatch <http://rdf.chemspider.com/370> .
\end{verbatim}
\end{tiny}

@@

The InChI is a unique identifier for (small) organic molecules, and has been previously used
to define unique IRIs for molecules~\cite{Bradley2009,Willighagen2011}. While IRIs are theoretically unlimited in length,
in practice web browsers and servers do limit the length of IRIs. Virtuoso is, unfortunately,
a system which supports only IRIs of up to a certain length. Therefore, only for smaller molecules
InChI-based links are created. Almost 1.3 million links are created, looking like:

@@

The Linked Data server used by Bio2RDF has been reconfigured for ChEMBL to provide URL
based services for standard URI resolution, along with text and link searches~\cite{Ansell2011}. 
The configuration files along with build scripts for the ChEMBL-RDF Linked Data server are available on GitHub~\cite{WebAppGitHub}. 
It proxies the standard item URIs by translating URLs between those requested by users and the URIs that are available
in SPARQL endpoints. For example, if the ChEMBL web application is running on the users
local machine, e.g. \url{http://localhost:8080/chembl/}, then a request for the
article with identifier ``a31863'', \url{http://localhost:8080/-chembl/article/a31863},
will be resolved from the database using the full original URI,
\url{http://data.kasabi.com/-dataset/chembl-rdf/13/activity/a31863}. If the
user requested an RDF document using content negotiation, the original URIs will be unchanged,
however, if the user requested an HTML document, the results will contain both the
original RDF triples, represented using RDFa, with links that resolve using the users local machine.

The links services enables the ChEMBL application to derive both forward links, originating in
ChEMBL, e.g. \url{http://localhost:8080/chembl/linkstonamespace/-targetns/originalns:identifier},
and backward links, originating in other databases, such as LODD, Bio2RDF, e.g.
\url{http://bio2rdf.org/linksns/targetns/originalns:ident-ifier}, or Chem2Bio2RDF.
These services are vital to efficiently navigate the Linked Data web, as it is both
impractical and inefficient to require users to crawl the web before they can discover all of the
relevant resources. These services are currently only supplied as web services from ChEMBL and
Bio2RDF, but it is hoped that similar services will be provided by other scientific Linked Data
providers in the future. Datasets that are available in SPARQL endpoints can be queried for
links efficiently using templated queries in the ChEMBL web application.

@@

While we show here that the triple version of the ChEMBL data is very useful, the current triples are by no means static: ChEMBL-RDF
will keep evolve, following new developments in the Linked Open Data world. For example, it is likely that over time we will link out
to more Linked Data resources. More importantly, we wish to adopt more common ontologies, of which the BioAssay Ontology is planned
to be the first next to be adopted~\cite{Visser2011}. This should make the ChEMBL-RDF triples even more interoperable.

>>>>

Although many parts of the ChEMBL-RDF 13 dataset now use community based ontologies, some 
terms from the previous ad-hoc ontology are still used. These terms were created under the 
URI namespace \textit{http://rdf.farmbio.uu.se/chembl/onto/\#}, and are referenced here using the
prefix: \textit{chembl}.

@@

The core concept in the ChEMBL database is that of the biological activity. This
is the type of information that the ChEMBL database extracts from literature.
In ChEMBL-RDF activities are mostly represented using the custom ChEMBL-RDF vocabulary, where the
type, fields and links to other resources are using the \textit{chembl}
namespace. While ChEMBL provides both the original activities as found in literature
and standardized values allowing comparison between studies, the triples only
make the latter available. The CiTO ontology is used to point to the paper from
which the data was extracted.

@@

\begin{footnotesize}
\begin{verbatim}
mol:m41
 rdfs:label "ChEMBL406142" , 
   "Bis(3-[1 .... yl]-propionamide)" ;
 rdfs:subClassOf cheminf:CHEMINF_000000 ;
 cheminf:CHEMINF_000200
  m41:inchikey , m41:smiles , m41:inchi ;

@@

chemblid:ChEMBL406142
 owl:equivalentClass mol:m41 .
\end{verbatim}
\end{footnotesize}

@@

To integrate ChEMBL-RDF with other RDF versions of scientific datasets, we link out to various resources.
These links are shown in Figure~\ref{2}. Triples for compounds link out to ChemSpider 
using the complementary index and OpenMolecules RDF using InChI values. Protein links are given to Bio2RDF
using the Uniprot identifier. Literature references are directed to CrossRef and Bio2RDF 
using the DOIs and PubMed identifiers, respectively.

@@

%FIXME:
For protein links out are based on the FOOBAR for the species, the EC code for proteins,
as well as UniProt identifiers.

FIXME: should we link out to the new UniProt RDF? PA: Code is available on the OpenPHACTS branch to link to http://purl.uniprot.org/uniprot/

@@

\begin{tiny}
\begin{verbatim}
<http://data.kasabi.com/dataset/chembl-rdf/chemblid/ChEMBL324846>
 skos:exactMatch <http://rdf.chemspider.com/370> .
\end{verbatim}
\end{tiny}

@@

The InChI is a unique identifier for (small) organic molecules, and has been previously used
to define unique IRIs for molecules~\cite{Bradley2009,Willighagen2011}. While IRIs are theoretically unlimited in length,
in practice web browsers and servers limit the length of IRIs. Virtuoso is, unfortunately,
a system which supports only IRIs of up to a certain length. Therefore, InChI-based links are only created for smaller molecules. 
Almost 1.3 million links to OpenMolecules RDF were created in a similar manner to:

@@

The Linked Data server used by Bio2RDF has been reconfigured for ChEMBL to provide URL
based services for standard URI resolution, along with text and link searches~\cite{Ansell2011}. 
The configuration files along with build scripts for the ChEMBL-RDF Linked Data server are available on GitHub~\cite{WebAppGitHub}. 
It proxies the standard ChEMBL-RDF URIs by translating URLs between those requested by users and the URIs that are available
in SPARQL endpoints. For example, if the ChEMBL web application is running on the users
local machine, e.g. \url{http://localhost:8080/chembl/}, then a request for the
article with identifier ``a31863'', \url{http://localhost:8080/-chembl/article/a31863},
will be resolved from the database using the full original URI,
\url{http://data.kasabi.com/-dataset/chembl-rdf/13/activity/a31863}. If the
user requested an RDF document using content negotiation, the original URIs will be unchanged,
however, if the user requested an HTML document, the results will contain both the
original RDF triples, represented using RDFa, with links that resolve using the users local machine.

The links services enables the ChEMBL application to derive both forward links, originating in
ChEMBL, e.g. \url{http://localhost:8080/chembl/linkstonamespace/-targetns/originalns:identifier},
and backward links, originating in other databases, such as LODD, Bio2RDF,
\footnote{\url{http://bio2rdf.org/linksns/targetns/originalns:identifier}}, and Chem2Bio2RDF.
These services are vital to efficiently navigate the Linked Data web, as it is both
impractical and inefficient to require users to crawl the entire web before they can discover 
relevant resources. These services are currently only supplied as web services from ChEMBL and
Bio2RDF, but it is hoped that similar services will be provided by other scientific Linked Data
providers in the future. Datasets that are available in SPARQL endpoints can be queried for
links efficiently using simple queries as demonstrated in the ChEMBL web application.

@@

While we show here that the RDF version of the ChEMBL data is very useful, the current RDF triples are by no means static: ChEMBL-RDF
will keep evolve, following new developments in the Linked Open Data world. For example, it is likely that over time we will link out
to more Linked Data resources. More importantly, we wish to adopt more common ontologies, of which the BioAssay Ontology is planned
to be the next~\cite{Visser2011}. These steps will make the ChEMBL-RDF triples even more interoperable.
